---
title: "SS2864B Assignment 2"
author: "Ali Al-Musawi"
date: "01/02/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T)
```
## Question 1
In this question, we create a vector and a dataframe.
First, create a vector that holds the elements $-20^2, -19^2, ..., 19^2, 20^2$.
Next, create a dataframe that holds the table of values of $y = x^2$.
Then, save the vector to a file named "squares"and the dataframe to a file named "quadratic". Finally, read in the saved objects and check if they are equvalent to the previously created objects. 
```{r}
vector <- I((-20:20)^2)
df <- data.frame(-20:20, vector)
names(df) <- c("x", "x^2")
write(vector, file = "squares")
write.table(df, "quadratic")
vector1 <- scan("squares")
df1 <- read.table("quadratic")
vector1 == vector && df == df1
```
Running this code, we get "TRUE", which means they are equivalent.

\newpage

## Question 2
In this question, we find out for which elements of the sequences $a_n = \{2^1, 2^2, ..., 2^{15}\}$ and $b_n = \{1^3, 2^3, ..., 15^3\}$ we have that $a_i > b_i$.
```{r}
powers2 <- 2^(1:15)
cubics <- (1:15)^3
n <- length(powers2)
(1:n)[powers2 > cubics]
```
Running this code, we find out the $a_i>b_i, \forall i\in\{1, 10, 11, 12, 13, 14, 15\}.$

\newpage

## Question 3
In this question, we examine the differeces of the R functions **save** and **dump**.
```{r}
x <- 1:10
y <- as.factor(rep(c("M", "F"), 5))
z <- data.frame(x, y)
dump("x", "x.R")
dump("y", "y.R")
dump("z", "z.R")
save("x", file= "x.RData")
save("y", file= "y.RData")
save("z", file= "z.RData")
xx <- x
yy <- y
zz <- z
rm(list = c("x", "y", "z"))
source("x.R")
source("y.R")
source("z.R")
(x == xx) && (y == yy) && (z == zz)
rm(list = c("x", "y", "z"))
load("x.RData")
load("y.RData")
load("z.RData")
(x == xx) && (y == yy) && (z == zz)
```
Running this code and examining its results, we conclude with the following: both functions allow us to save objects created here. When importing the saved objects, there is no need to assign the saved objects to a new variable as they keep the original name, and we note that they are equivalent to the original objects. The function **dump** saves R objects in text (script) format, and to retrieve the object, a call to **source** is needed. On the other hand, the function **save** saves R objects in binary format, and to retrieve the object, a call to the function **load** is required. By convention, R binary objects are saved into **.RData** extension files, whereas R scripts are saved into **.R** extension files.

\newpage

## Question 4
In this question, we categorize BMI data according to the table in this link: [Wikipedia: BMI](https://en.wikipedia.org/wiki/Body_mass_index). We first categorize the data into 11 intervals, find out about the frequency of records belonging to each interval, then make a barplot of our findings. 
```{r}
library(MASS)
data(Pima.te)
fac1 <- cut(x = Pima.te$bmi, breaks = c(0, 15, 16, 18.5, 25, 30, 35, 40, 45, 50, 60, Inf))
levels(fac1)[1] <- "Very severely underweight"
levels(fac1)[2] <- "Severely underweight"
levels(fac1)[3] <- "Underweight"
levels(fac1)[4] <- "Normal (healthy weight)"
levels(fac1)[5] <- "Overweight"
levels(fac1)[6] <- "Obese Class I (Moderately obese)"
levels(fac1)[7] <- "Obese Class II (Severely obese)"
levels(fac1)[8] <- "Obese Class III (Very severely obese)"
levels(fac1)[9] <- "Obese Class IV (Morbidly obese)" 
levels(fac1)[10] <- "Obese Class V (Super obese)"
levels(fac1)[11] <- "Obese Class VI (Hyper obese)"
table(fac1)
levels(fac1) <- 1:11
plot(fac1, main = "Frequency of BMI Categories Recorded")
```
Next, we repeat the same process, but instead we restrict it to 3 intervals.
```{r}
fac2 <- cut(x = Pima.te$bmi, breaks = c(0, 18.5, 25, Inf))
levels(fac2)[1] <- "Underweight"
levels(fac2)[2] <- "Normal (healthy weight)"
levels(fac2)[3] <- "Overweight"
table(fac2)
plot(fac2, main = "Frequency of BMI Categories Recorded")
```

\newpage

## Question 5
In this question, we examine the relationship between *temperature* and *pressure* in the **pressure** dataset. First, we produce a scatterplot of the variables.
```{r}
data("pressure")
plot(x = pressure$temperature, y = pressure$pressure)
```
Based on this graph, we conclude the relationship between the two variables is non-linear.
Next, we assume that the relationship between *pressure* $P$ and *temperature* $T$ is given by $P = (0.168+0.007T)^\frac{20}{3}$. We compute the residuals of this function and the observed values, and produce a normal QQ plot of the residuals.
```{r}
residuals <- with(pressure, pressure-(0.168+0.007*temperature)^(20/3))
qqnorm(residuals)
qqline(residuals)
```
Based on the plot, we see that the residuals are skewed and not normally distributed.  
  
Now, we repeat the process above, but instead we transform $P$ to $P^\frac{3}{20}$. The following is a scatterplot of the data.
```{r}
plot(x = pressure$temperature, y = I(pressure$pressure^(3/20)))
```
The relationship here appears to be linear.
Now, we compute residuals associated with the transformed pressure to examine their distribution:
```{r}
residuals <- with(pressure, pressure^(20/3)-(0.168+0.007*temperature)^(20/3))
qqnorm(residuals)
qqline(residuals)
```
Based on this graph, we see that the residuals are slightly skewed.


\newpage

## Question 6
In this question, we continue working with the previous dataset. We fit a curve to $P$ as a function of $T$:
```{r}
plot(x = pressure$temperature, y = pressure$pressure,
     xlab = "Temperature", ylab = "Pressure",
     main = "Temperature VS Pressure")
legend(x = "bottomright", legend = c("Pressure"), xpd = T)
curve((0.168+0.007*x)^(20/3), from = 0, to = 400, add = TRUE)
```

Next, we fit a line to the transformed pressure $P^\frac{3}{20}$ as a function of $T$:
```{r}
plot(x = pressure$temperature, y = I(pressure$pressure^(3/20)), 
     xlab = "Temperature", ylab = "Pressure^(3/20)", 
      main = "Temperature VS Transformed Pressure")
legend(x = "bottomright", legend = c("Transformed Pressure"), xpd = T)
fit <- lm(I(pressure^(3/20))~temperature, data = pressure)
abline(fit)
```

Now, we produce a $2 \times 1$ plot of the two fits above:
```{r}
par(mfrow=c(2,1))
plot(x = pressure$temperature, y = pressure$pressure,
     xlab = "Temperature", ylab = "Pressure",
     main = "Temperature VS Pressure")
curve((0.168+0.007*x)^(20/3), from = 0, to = 400, add = TRUE)
legend(x = "bottomright", legend = c("Pressure"), xpd = T)
plot(x = pressure$temperature, y = I(pressure$pressure^(3/20)), 
     xlab = "Temperature", ylab = "Pressure^(3/20)", 
     main = "Temperature VS Transformed Pressure")
abline(fit)
legend(x = "bottomright", legend = c("Transformed Pressure"), xpd = T)

```
Finally, we produce a $1 \times 2$ plot of the two fits above:
```{r}
par(mfrow=c(1,2))
plot(x = pressure$temperature, y = pressure$pressure,
     xlab = "Temperature", ylab = "Pressure",
     main = "Temperature VS Pressure")
curve((0.168+0.007*x)^(20/3), from = 0, to = 400, add = TRUE)
legend(x = "bottomright", legend = c("Pressure"), xpd = T)
plot(x = pressure$temperature, y = I(pressure$pressure^(3/20)), 
     xlab = "Temperature", ylab = "Pressure^(3/20)", 
     main = "Temperature VS Transformed Pressure")
abline(fit)
legend(x = "bottomright", legend = c("Transformed Pressure"), xpd = T)
```
Please note that the points run through the legend, but this problem is due to knitting with R markdown. If the code is used in an R script, this problem does not occur.  Please review the accompanying R script for the source code.