---
title: "SS2864B Assignment 5"
author: "Ali Al-Musawi"
date: "13/03/2020"
header-includes:
   - \usepackage{amsmath}
output: pdf_document
latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1
```{r}
# define the given function
f1 = function(x) {
  return ((cos(x))^2)
} 

# define a CI function
ci.t.95 = function(x) {
  # compute lower and upper bounds 
  lower = mean(x) - 1.96*sd(x)/sqrt(length(x))
  upper = mean(x) - 1.96*sd(x)/sqrt(length(x))
  # construct a list object
  result = list(Lower.Bound = lower, Mean = mean(x), Upper.Bound = upper)
  return (result)
}
```

### (a)
```{r}
curve(f1, from = 0, to = 2*pi, xname = 'x',
      ylab = 'f(x)', main = ' A plot of [cos(x)]^2',
      col = 'blue')
grid(5, lwd = 2)
```


### (b)
```{r}
# set seed for reproducibility
set.seed(123)
# define a random uniform variable on [0, 2pi]
U = runif(n = 1e6, min= 0, max = 2*pi)
# perform MC integration
I1 = mean(f1(U))*(2*pi)
# compute a 95% ci
ci = ci.t.95(f1(U)*(2*pi))
# display ci
ci
```

### (c)
We use some identities to simplify the square of cosine before integrating it:
$$\cos^2(x) = 1 - \sin^2(x) \text{ (Pythagorean Identity)}$$
$$\sin^2(x) = \cos^2(x) - \cos(2x)   \text{ (Cosine Double Angle Identity)}$$
$$\therefore 2\cos^2(x) = 1 - \cos(2x) \text{ (Combine Two Identities)}$$
$$I =\int_{0}^{2\pi} f(x) dx = \int_{0}^{2\pi} \cos^2(x) dx = \frac{1}{2}\int_{0}^{2\pi} 1 - \cos(2x)  dx $$
$$I = \frac{1}{2} \left( \left. x - \frac{1}{2} \sin(2x) \right) \right\rvert_{0}^{2\pi}$$
$$\therefore I = \frac{1}{2} \left( 2\pi\right) = \pi $$
Based on the output in part **b**, indeed, the CI covers the true value if we assume  the true value = 3.14.

## Question 2
### (a)
```{r}
set.seed(222)
# Generate 100 x_i such that X~N(10, 4)
x = rnorm(n = 100, mean = 10, sd = 2)
# Generate summary statistics of the x_i's
min(x)
max(x)
mean(x)
sd(x)
```
### (b)
```{r}
# Generate 50000 different samples and calculate their means
out = replicate(50000, mean(sample(x, replace=TRUE)))
# Generate summary statistics of the means
min(out)
max(out)
mean(out)
sd(out)
```
### (c)
```{r}
# Construct a histogram with double the number of bins
hist(x = out, freq = F, xlab = 'Boostrapped Means',
     main = 'Sampling Distribution of Sample Mean',
     breaks=nclass.Sturges(out)*2)
# Add a density line
lines(density(out, adjust=2), col=2, lwd=2)
# Add a verticle line at the centre of the distribution
abline(v = median(out), col = 4)
```
This output looks like a translated normal distribution centered at the true sample mean.

### (d)
```{r}
# Construct a histogram with double the number of bins
hist(x = out-mean(x), freq = F, xlab = 'Shifted Boostrapped Means',
     main = 'Sampling Distribution of Sample Mean (Shifted)',
     breaks=nclass.Sturges(out)*2)
# Add a density line to the histogram
lines(density(out - mean(x), adjust=2), col=2, lwd=2)
# Add a verticle line at the centre of the distribution
abline(v = median(out - mean(x)), col = 6)
# Add the theoretical distribution
lines(density(out - 10, adjust=2), col=4, lwd=2, lty = "dashed")
# Add a verticle line at the centre of the theoretical distribution
abline(v = median(out - 10), col = 6)
# Add a legend
legend("topright", col = c(2, 4), lty = c(1, 2), legend = 
         c('Boostrap Distribution', 'Theoretical Distribution'))
```
It looks like the boostrap distribution is identical to theoretical distribution except that their centres are shifted (translated). This shift may be called an 'error'. Note that the distributions are equivariant. 

### (e)
We will omit the code from the output since it is identical to the last parts. 
** Round 2 **
```{r, echo = FALSE}
set.seed(223)
# Generate 100 x_i such that X~N(10, 4)
x = rnorm(n = 100, mean = 10, sd = 2)
# Generate summary statistics of the x_i's
print('summary statistics of the sample: ')
print('min:')
min(x)
print('max:')
max(x)
print('mean:')
mean(x)
print('sd:')
sd(x)
```

```{r, echo = FALSE}
# Generate 50000 different samples and calculate their means
out = replicate(50000, mean(sample(x, replace=TRUE)))
# Generate summary statistics of the means
print('summary statistics of the means (boostrapped data): ')
print('min:')
min(out)
print('max:')
max(out)
print('mean:')
mean(out)
print('sd:')
sd(out)
```

```{r, echo = FALSE}
# Construct a histogram with double the number of bins
hist(x = out, freq = F, xlab = 'Boostrapped Means',
     main = 'Sampling Distribution of Sample Mean',
     breaks=nclass.Sturges(out)*2)
# Add a density line
lines(density(out, adjust=2), col=2, lwd=2)
# Add a verticle line at the centre of the distribution
abline(v = median(out), col = 4)
```

```{r, echo = FALSE}
# Construct a histogram with double the number of bins
hist(x = out-mean(x), freq = F, xlab = 'Shifted Boostrapped Means',
     main = 'Sampling Distribution of Sample Mean (Shifted)',
     breaks=nclass.Sturges(out)*2)
# Add a density line to the histogram
lines(density(out - mean(x), adjust=2), col=2, lwd=2)
# Add a verticle line at the centre of the distribution
abline(v = median(out - mean(x)), col = 6)
# Add the theoretical distribution
lines(density(out - 10, adjust=2), col=4, lwd=2, lty = "dashed")
# Add a verticle line at the centre of the theoretical distribution
abline(v = median(out - 10), col = 6)
# Add a legend
legend("topright", col = c(2, 4), lty = c(1, 2), legend = 
         c('Boostrap Distribution', 'Theoretical Distribution'))
```

Note in general the results are similar, except that this time, the 'error' is larger.

** Round 3 **
```{r, echo = FALSE}
set.seed(231)
# Generate 100 x_i such that X~N(10, 4)
x = rnorm(n = 100, mean = 10, sd = 2)
# Generate summary statistics of the x_i's
print('summary statistics of the sample: ')
print('min:')
min(x)
print('max:')
max(x)
print('mean:')
mean(x)
print('sd:')
sd(x)
```

```{r, echo = FALSE}
# Generate 50000 different samples and calculate their means
out = replicate(50000, mean(sample(x, replace=TRUE)))
# Generate summary statistics of the means
print('summary statistics of the means (boostrapped data): ')
print('min:')
min(out)
print('max:')
max(out)
print('mean:')
mean(out)
print('sd:')
sd(out)
```

```{r, echo = FALSE}
# Construct a histogram with double the number of bins
hist(x = out, freq = F, xlab = 'Boostrapped Means',
     main = 'Sampling Distribution of Sample Mean',
     breaks=nclass.Sturges(out)*2)
# Add a density line
lines(density(out, adjust=2), col=2, lwd=2)
# Add a verticle line at the centre of the distribution
abline(v = median(out), col = 4)
```

```{r, echo = FALSE}
# Construct a histogram with double the number of bins
hist(x = out-mean(x), freq = F, xlab = 'Shifted Boostrapped Means',
     main = 'Sampling Distribution of Sample Mean (Shifted)',
     breaks=nclass.Sturges(out)*2)
# Add a density line to the histogram
lines(density(out - mean(x), adjust=2), col=2, lwd=2)
# Add a verticle line at the centre of the distribution
abline(v = median(out - mean(x)), col = 6)
# Add the theoretical distribution
lines(density(out - 10, adjust=2), col=4, lwd=2, lty = "dashed")
# Add a verticle line at the centre of the theoretical distribution
abline(v = median(out - 10), col = 6)
# Add a legend
legend("topright", col = c(2, 4), lty = c(1, 2), legend = 
         c('Boostrap Distribution', 'Theoretical Distribution'))
```
Again, we get similar results, and this time the 'error' has shrunk.

## Question 3

### (a)
```{r}
my.obj = function(theta, x) {
  # We allow theta to be a vector. 
  # If dim(theta) > 1, we return a vector of dim(theta)
  #   and we re-use the data in x for all theta given
  # Else, we return a scalar
  if (length(theta) > 1) {
    x.mat = matrix(rep(x, times = length(theta)), nrow = length(x))
    theta.mat = matrix(rep(theta, each = length(x)), nrow = length(x))
    result = colSums(abs(x.mat - theta.mat))
  }
  else {
    result = sum(abs(x - theta))
  }
  return(result)
}
```

### (b)
```{r}
# define a function that minimizes my.obj with respect to theta
my.minimizer1 = function(x, interval = c(min(x), max(x))) {
  o = optimize(my.obj, interval = c(min(x), max(x)), x)
  return(o$minimum)
}

# test the function on the given data
data = c(3, 7, 9, 12, 15, 18, 21)
my.minimizer1(data)
median(data)

```
Our function works perfectly!

### (c)
```{r}
# define a function that minimizes my.obj with respect to theta
my.minimizer2 = function(x, start = mean(x)) {
  o = nlminb(start = start, objective = my.obj, x = x)
  return(o$par)
}

# test the function on the given data
my.minimizer2(data)
median(data)

```
This function works just as good too.

### (d)
```{r}
# create test data
data = c(1, 3, 7, 9, 12, 15, 18, 21)
# test the functions from b and c
tb1 = my.minimizer1(data, interval = c(min(data) - 2, max(data) + 2))
tb2 = my.minimizer1(data, interval = c(min(data) - 4, max(data) + 4))
tb3 = my.minimizer1(data, interval = c(min(data) - 6, max(data) + 6))
tc1 = my.minimizer2(data, start = min(data))
tc2 = my.minimizer2(data, start = max(data))
tc3 = my.minimizer2(data, start = mean(data)/2)
tb1
tb2
tb3
tc1
tc2
tc3
median(data)
```
It seems that both functions fail to converge to the minimizer.
```{r}
library('MASS')
# evaluate theta on interval [1, 21]
theta = seq(min(data), max(data), length.out = 1e5)
# evaluate f on theta and the given data
f = my.obj(theta, data)
# set up a grid and plotting area
plot(x = theta, y = f, type = 'n', main = 'A plot f(theta|x), x in {2n+1}, n in N',
     xlab = 'theta', ylab = 'f(theta|x)', col = 4)
grid(5, lwd = 2)
# create a smooth plot of the data
sm = smooth.spline(theta, f, spar=1)
# add the smooth plot
lines(sm, col = 4)
abline(v = median(data), col = 3, lty = 2)
abline(h = my.obj(median(data), data), col = 3, lty = 2)
```
We find that indeed, the value of $\theta$ that minimizes $f(\theta|x)$ is the median of $x$. In this case, it is 10.5. 

## Question 4
We will omit the code from showing on the output due to its length.
```{r, echo=FALSE}
"huron" <- 
structure(.Data = c(581.55999999999995, 581.54999999999995, 581.34000000000003, 580.84000000000003, 
	580.33000000000004, 580.35000000000002, 579.87, 580.49000000000001, 
	579.90999999999997, 580.07000000000005, 580.90999999999997, 581.10000000000002,
	579.72000000000003, 580.32000000000005, 580.48000000000002, 580.38, 
	581.86000000000001, 580.97000000000003, 580.79999999999995, 579.78999999999996,
	580.38999999999999, 580.41999999999996, 580.82000000000005, 581.39999999999998,
	581.32000000000005, 581.44000000000005, 581.67999999999995, 581.16999999999996,
	580.52999999999997, 580.00999999999999, 579.90999999999997, 579.13999999999999,
	579.15999999999997, 579.54999999999995, 579.66999999999996, 578.44000000000005,
	578.24000000000001, 579.10000000000002, 579.09000000000003, 579.35000000000002,
	578.82000000000005, 579.32000000000005, 579.00999999999999, 579., 
	579.79999999999995, 579.83000000000004, 579.72000000000003, 579.88999999999999,
	580.00999999999999, 579.37, 578.69000000000005, 578.19000000000005, 
	578.66999999999996, 579.54999999999995, 578.91999999999996, 578.09000000000003,
	579.37, 580.13, 580.13999999999999, 579.50999999999999, 579.24000000000001, 
	578.65999999999997, 578.86000000000001, 578.04999999999995, 577.78999999999996,
	576.75, 576.75, 577.82000000000005, 578.63999999999999, 580.58000000000004, 
	579.48000000000002, 577.38, 576.89999999999998, 576.94000000000005, 
	576.24000000000001, 576.84000000000003, 576.85000000000002, 576.89999999999998,
	577.78999999999996, 578.17999999999995, 577.50999999999999, 577.23000000000002,
	578.41999999999996, 579.61000000000001, 579.04999999999995, 579.25999999999999,
	579.22000000000003, 579.38, 579.10000000000002, 577.95000000000005, 578.12, 579.75,
	580.85000000000002, 580.40999999999997, 579.96000000000004, 579.61000000000001,
	578.75999999999999, 578.17999999999995, 577.21000000000004, 577.13, 
	579.10000000000002, 578.25, 577.90999999999997, 576.88999999999999, 
	575.96000000000004, 576.79999999999995, 577.67999999999995, 578.38, 
	578.51999999999998, 579.74000000000001, 579.30999999999995, 579.88999999999999,
	579.96000000000004, 580.98000000000002, 581.03999999999996, 580.49000000000001,
	580.51999999999998, 578.57000000000005, 578.96000000000004, 579.94000000000005,
	579.76999999999998, 579.44000000000005, 578.97000000000003, 580.08000000000004,
	580.23000000000002, 580.75, 581.26999999999998)
, class = c("ts", "numeric")
, tsp = c(1860., 1986., 1.)
, title = "Lake Huron, mean level, July, 1860-1986"
)
```
### (a)
```{r}
mean(huron)
sd(huron)
min(huron)
max(huron)
boxplot(huron, main = "Huron Lake Mean Water Level Over 1860-1986")
ts.plot(huron, main = "Huron Lake Mean Water Level Over 1860-1986", ylab = "Mean Water Level")
```
### (b)
```{r}
# define the given likelihood function
log.likelihood=function(par, x){
  n=length(x)
  v=x[1]^2
  for (i in 2:n)
    v=v+(x[i]-par[1]*x[i-1])^2
  return(v/par[2]+n*log(par[2]))
}
# define huron2
huron2 = huron - mean(huron)
# do optimization
my.par = nlminb(start = c(-0.1, 540), objective = log.likelihood, x = huron2)
# extract the parameters
par = my.par$par
# display the parameters
par
```
### (c)
```{r}
# predict the time series data using the first entry in huron2
pred.huron2 = huron2
for (i in 2:length(huron2)) {
  pred.huron2[i] = par[1]* huron2[i-1] 
}
# plot the predicted and actual time series
ts.plot(huron2, pred.huron2, col = c('red', 'blue'), 
        main = "Huron Lake Shifted Mean Water Level Over 1860-1986",
        ylab = "Shifted Mean Water Level")
# add a legend and a grid
grid(5, lwd = 2)
legend("bottomleft", col = c('red', 'blue'), lty = 1, legend = c("Actual", "Predicted"))

```
Our model captures the fluctuations of the time series data; however, it does not match the original data and we can see it is off by a constant.
